\section{Algorithm's Correctness} 
We will prove that if the original locking protocol is 
conflict-serializable then our algorithm is conflict-serializable.

Let $\pi$ be an execution of our optimistic automation on a 
sequential algorithm. We will construct an execution $\pi_{LP}$ 
which is an execution following the original locking protocol. 
We will prove that both executions are conflict-equivalent. 
Since any execution of the original locking protocol
is conflict-serializable, then $\pi$ is conflict-serializable. 

Let $p_1,p_2,\ldots,p_n$ be the operations $\in\pi$ ordered by the 
order of execution of the first step of a successful \readSet 
validation. (If some operation does not have such point we omit it).
Let $\pi_{LP} = \pi_{lp1},\pi_{1},\ldots,\pi_{lpi},\pi_{i}$ where 
$\pi_{lpi}$ is a $p_i$-only execution of original locking protocol 
until $p_i$ holds locks only on the local variable locked
in the validation phase of $p_i \in \pi$, and $\pi_i$ is
the interval of $\pi$ starting from the return from the validation of
$pi$ until the first step of the successful \readSet validation of 
$p_{i+1}$ that includes only the operations by $\{p_1,\ldots,p_i\}$.
In other words, we replace the read-phase and validation phase with 
an execution of the original locking protocol, 
taking place at the point just before the \readSet validation starts. 

%TODO connect the requirement on the locking protocol to the construction.
\begin{lemma}
The construction of $\pi_{LP}$ is feasible.  
\end{lemma}
\begin{proof}
Proof by induction on $p_1,p_2,\ldots,p_n$. Base case is immediate. 

Let $\pi' = \pi_{lp1},\pi_{1},\ldots,\pi_{lpk-1},\pi_{k-1}$ be the feasible
construction so far, and let $p_{k}$ be the next operation to be 
added. 

Assume by contradiction that $\pi'\cdot\pi_{lpk}\cdot\pi_{k}$ 
cannot be constructed, thus, some object $v$ that $p_{k}$ locks 
in $\pi_{lpk}$ is already locked 
by $p_j \in \{p_1,p_2,\ldots,p_{k-1}\}$ in
the last configuration of $\pi'$. 
If $p_j$ locked $v$ before $p_k$ read $v$ for the first time, 
then $v$ was locked during the read phase of $p_k$, 
in contradiction to $p_k$ reaching its validation. 
Otherwise, $p_j$ locked $v$ after $p_k$ read $v$. 
If $v$ is still locked during the validation of $p_k$ then 
the validation will fail, contradiction. Alternatively, $v$ 
was unlocked by $p_j$ before $p_k$ validated $v$, 
its version incremented to a version bigger 
than the local version read by $p_k$, 
contradicting the successful validation of $p_k$.  
\end{proof}

\begin{lemma}
$\pi_{LP}$ is conflict-equivalent to $\pi$
\end{lemma}
\begin{proof}
Each operation performs a double collect on all the values it reads. 
The first collect is the read phase and the second is the \readSet 
validation of the validation phase. Since validation was successful, 
both collect are identical, meaning that the values of the \readSet
do not change from the return of the last read of the read phase,
until the first read of the \readSet validation. Therefore, executing 
the original locking of $p_k$ after $\pi' =
\pi_{lp1},\pi_{1},\ldots,\pi_{lpk-1},\pi_{k-1}$ is conflict-equivalent
to the original read phase. The read-write phase remains unchanged, 
maintaining conflict-equivalence to $\pi$.
\end{proof}