%
\documentclass{sig-alternate}

%
% the following standard packages may be helpful, but are not required
%

%\usepackage{SIunits}            % typset units correctly
%\usepackage{courier}            % standard fixed width font
%\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                  % format URLs
%\usepackage{listings}          % format code
%\usepackage{enumitem}      % adjust spacing in enums
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography

\usepackage{amsmath}

\usepackage{xspace}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{subcaption}
\usepackage{xcolor}
\usepackage{tikz}
\usepackage{pgfplots}

\input{plotsStyle}

\input{macros}

% Add page numbers, remove copyright box.  For submitted version only.
\pagenumbering{arabic}
\makeatletter
\def\@copyrightspace{\relax}
\makeatother

\begin{document}

%\special{papersize=8.5in,11in}
%\setlength{\pdfpageheight}{\paperheight}
%\setlength{\pdfpagewidth}{\paperwidth}

%\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
%\copyrightyear{20yy}
%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
%\doi{nnnnnnn.nnnnnnn}

\title{Towards Automatic Lock Removal\\ for Scalable Synchronization}
\author{
Maya Arbel\footnotemark[1] \footnotemark[2] \hspace{0.3in} 
Guy Golan Gueta\footnotemark[1]  \hspace{0.3in}
Eshcar Hillel\footnotemark[1]  \hspace{0.3in}
Idit Keidar\footnotemark[1] \footnotemark[2]\\
\\
\large{
\footnotemark[1]Yahoo Labs, Haifa, Israel \hspace{0.3in} 
\footnotemark[2]The Technion, Haifa, Israel
}
}


\maketitle

\begin{abstract}
We present a \emph{code transformation} for concurrent data structures,
which increases their scalability without sacrificing correctness.
Our transformation takes lock-based code, and replaces some of the
locking steps therein with optimistic synchronization, in order to reduce contention. The main idea is to
have each operation perform an optimistic traversal of the data structure
as long as no shared memory locations are updated, and then proceed with
pessimistic code. The transformed code inherits essential
properties of the original one, including linearizability, serializability,
and deadlock freedom.
%It reduces contention, both on locks and on access to shared memory locations.
%When applying our transformation to
%hand-over-hand locking solutions, we obtain significantly superior scalability.

Our work complements existing pessimistic transformations that make
sequential code thread-safe by adding locks.
In essence, we provide a way to optimize such transformations by reducing
synchronization bottlenecks (for example, locking the root of a tree).
The resulting code scales well and significantly outperforms
pessimistic approaches. We further compare our synthesized code to state-of-the-art
data structures implemented by experts.
We find that its performance is comparable %, and sometimes even superior,
to that achieved by the custom-tailored implementations.
Our work thus shows the promise that automated approaches
bear for overcoming the difficulty involved in manually
hand-crafting concurrent data structures.

\end{abstract}


\input{introduction}
%\input{preliminariesB}
\input{algorithmB}
\input{proof2}
\input{evaluation2}
\input{related}
\input{discussion}
%\section{Appendix Title}

%This is the text of the appendix, if you need one.

%\acks

\bibliography{myRef}
\bibliographystyle{abbrv}
%

\appendix
\input{proof-appendix}


\end{document}
